---
title: "Sobel with an ODC"
author: "Eva Newby"
format: html
editor_options: 
  chunk_output_type: console
execute: 
  warning: false
  message: false
---

```{r}
# Load packages
library(tidyverse)
library(deSolve)
library(sensitivity)
library(here)
```

```{r}
# read in forest growth function
source(here("forest_growth.R"))
```

## Run the model for 300 years
```{r}
# ODE needs ALL parameters in a single list
# initial population and times for which we want output
Cinitial <- c(C = 10)

# gets results for 300 years (evaluating every year)
simtimes <- seq(from = 1, to = 300)
parms <- list(r = 0.01, K = 250, g = 2, thresh = 50)

result <- ode(y = Cinitial, time = simtimes, func = dforestgrowth, parms = parms)
result <- as.data.frame(result)
```

## Graph results for 300 years
```{r}
ggplot(result, aes(time, C))+
  geom_line() +
  geom_hline(yintercept = 50, col = "cyan") +
  labs(y = "Forest Size (C)", x = "Time (years)", title = "Forest Growth Model over 300 years")+
  theme_minimal()
```

## Graph Explanation

The forest begins to grow slowly, but then accelerates once it reaches a threshold around 170 years. This information is important when thinking about forest management, and that harvesting any sort of trees and/or protection from environmental disturbances, such as wildfire, should be emphasized in the first 170 years. 

## Run a Sobol Sensitivity Analysis
```{r}
# Create two independent parameter sets
# first one
np <- 2000
K <- rnorm(mean = 250, sd = 25, n = np)
r <- rnorm(mean = 0.01, sd = 0.001, n = np)
g <- rnorm(mean = 2, sd =0.2 , n = np)
thresh <- rnorm(mean = 50, sd = 5, n = np)

X1 <- cbind.data.frame(r = r, K = K, g = g, thresh = thresh)

# second set
K <- rnorm(mean = 250, sd = 25, n = np)
r <- rnorm(mean = 0.01, sd = 0.001, n = np)
g <- rnorm(mean = 2, sd =0.2 , n = np)
thresh <- rnorm(mean = 50, sd = 5, n = np)

X2 <- cbind.data.frame(r = r, K = K, g = g, thresh = thresh)

# Fix any negative values
X1 <- X1 %>% map_df(pmax, 0.0)
X2 <- X2 %>% map_df(pmax, 0.0)

# Set up sobol saltelli sampling
sens_C <- sobolSalt(model = NULL, 
                    X1, 
                    X2, 
                    nboot = 300) # bootstrapping for confidence interval estimation

# define and add names
colnames(sens_C$X) <- c("r", "K", "g", "thresh")

# Define ODE model evaluation function
# returns the maximum forest size
max_C <- function(params) {
  parms <- list(
    r = params[1],
    K = params[2],
    g = params[3],
    thresh = params[4]
  )
  Cinitial <- c(C = 10)
  simtimes <- seq(from = 1, to = 300)
  result <- ode(y = Cinitial, time = simtimes, func = dforestgrowth, parms = parms)
  result <- as.data.frame(result)
  return(max(result$C))
}

# Run model on parameter samples
y <- apply(sens_C$X, 1, max_C)

# Compute and print sobol indices
sens_C <- tell(sens_C, y)

print("First-order Sobol indices (S):")
print(sens_C$S$original)

print("Total-effect Sobol indices (T):")
print(sens_C$T$original)
```

## Graph Box Plot of Maximum Forest Size
```{r}
# Get maximum forest size values (max values of c from each ODE run)
max_forest <- sens_C$y

# Box plot of max forest size with varied parameters
ggplot() + 
  geom_boxplot(aes(y = max_forest, x = ""),
               fill = "cyan", color = "black", outlier.color = "red", outlier.shape = 16) +
  labs(title = "Maximum Forest Size Sensitivity Analysis",
       y = "Maximum Forest Size (C)")+
  theme_minimal()
```

## Explanation of Results
The forest model shows variability in maximum size, generally ranging from ~165 to 200. Extreme values, such as <140 or >230, are possible, highlighting its sensitivity to the parameter inputs. This information is important to prioritize adaptive management and facilite further ecological research into reducing uncertainty for the parameters, such as with the disturbance threshold.

